CommonMind Consortium: DLPFC

### 2020-10-06 combined_analysis/combine_all_DLPFC.Rmd
Biol Psych freeze

}) 
source("../common_functions/common_functions.R")

@@

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
df_cca2 = df_cca[with(df_cca, v1!=v2),]

rstat_nodes <- data.frame(name = unique(df_cca2$v1))

@@

@@

@@

```{r test.multtest, eval=FALSE}

library(irlba)
library(pinnacle)
library(poolR)
library(matrixStats)

i = with(res_meta_zenith, (Resource == 'GO') & (Coef == "DxSCZ"))
df = res_meta_zenith[i,]
df$qvalue = qvalue(df$PValue)$qvalue
# df$p.adjust = p.adjust(df$PValue, method="bon" )
df$p.adjust = pmin(1, 2805 * df$PValue)

M = recodeToSparseMatrix(gs.GO)
M = M[rownames(M) %in% trim_ensembl_ids(rownames(df_meta_SCZ)),]
keep = (colSums(M) >= 10) & (colnames(M) %in% rownames(df) )
M = M[,keep]

m = colMeans(M)
s = apply(M, 2, var) * nrow(M) / (nrow(M) + 1)

dcmp = irlba( M, nv = 500, center=m, scale=sqrt(s), right_only=FALSE)
ev = dcmp$d^2 / nrow(M)

plot(cumsum(ev) / ncol(M))

df_unsort = df[colnames(M),]
edcmp = list(vectors = dcmp$v, values = dcmp$d^2 / nrow(M))

res = p.corr.adjust( df_unsort$PValue, dcmp = edcmp, method.sim="eigen", B = nrow(df_unsort)*4)

plot( cumsum(dcmp$d^2 / sum(dcmp$d^2)))

C = cor(as.matrix(M[,1:1000]))
evals = eigen(C)$values

dcmp = svd(scale(M) / sqrt(nrow(M)))
dcmp$d^2

plot(cumsum(dcmp$d^2 / sum(dcmp$d^2)))

meff(C, method='gao')

dcmp = svd(scale(M[,1:100]))

meff(eigen = dcmp$d^2 , method = "galwey")

@@

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$PValue)) * 1.05
  ggplot(df, aes(Set, -log10(PValue))) + geom_bar(stat="identity", fill="navy") + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~P)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax))
}

@@

# Concordance analysis
resConcord.Sex = plot_concordance(tab.de.lst.autosome, showGenes = tab.de.lst.autosome$Symbol[1:20])

@@

data.frame(pi1 = 1 - qvalue(df_meta[[coef]]$P)$pi0,
          `# DE` = sum(df_meta[[coef]]$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for Sex meta-analysis") %>% kable_styling(full_width=FALSE)
```

@@

@@

@@

```{r compare.separate, fig.width=5, fig.height=5}
df = merge( df_meta.separ[[1]], df_meta.separ[[2]], by="row.names")

fit = with(df, lm(logFC.y ~ logFC.x))

coef(summary(fit) ) %>%  kable(caption="logFC is larger in males") %>% kable_styling(full_width=FALSE)

 # compare logFC
tstat = with(df, (logFC.y - logFC.x) / (sqrt(se.y^2 + se.x^2)))

qqt(tstat)
abline(0, 1, col="red")
p = 2*pt(abs(tstat), df=778, lower.tail=FALSE)
table(p.adjust(p) < 0.05)
```

>>>>

}) 
source("../common_functions/common_functions.R")   
# source("common_functions/common_functions.R")

@@

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
df_cca2 = df_cca[with(df_cca, v1!=v2),]
df_cca2$key = apply(df_cca2, 1, function(x){
  x = sort(c(x['v1'], x['v2']))
  paste(x[1], x[2])
  }) 
df_cca2 = df_cca2[!duplicated(df_cca2$key),]

rstat_nodes <- data.frame(name = unique(c(df_cca2$v1, df_cca2$v2)))

@@

```{r fraction.female}
df = xtabs(~Reported_Gender + Institution, METADATA)
df2 = apply(df, 2, function(x) x / sum(x))

plotPercentBars(data.frame(t(df2))) + scale_fill_manual(values=c("red", "blue")) + ylab("Fraction of samples") + xlab("Institution")
```

@@

  plot_grid(fig1, fig2, nrow=1, rel_widths=c(.8, 1))
})

@@

@@

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$FDR)) * 1.05
  ggplot(df, aes(Set, -log10(FDR))) + geom_bar(stat="identity", fill="navy") + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~FDR)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed")
}

@@

# Concordance analysis
df = df_meta[[coef]][df_meta[[coef]]$Chrom %in% 1:22,]
resConcord.Sex = plot_concordance(tab.de.lst.autosome, showGenes = df$Symbol[1:20])

@@

data.frame(pi1 = 1 - qvalue(df_meta[[coef]]$P)$pi0,
          `# DE` = sum(df_meta[[coef]]$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for Sex meta-analysis") %>% kable_styling(full_width=FALSE)

data.frame(pi1 = 1 - qvalue(df$P)$pi0,
          `# DE` = sum(df$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for Sex meta-analysis - autosomes") %>% kable_styling(full_width=FALSE)
```

@@

# Comparison with GTEx
```{r gtex.sex, fig.height=6, fig.width=10}

# read beta and se from GTEx
df_gtex.beta = read.table(synGet("syn22891761")$path, header=TRUE, row.names=1)
df_gtex.se = read.table(synGet("syn22891767")$path, header=TRUE, row.names=1)
df_gtex.lfsr = read.table(synGet("syn22891755")$path, header=TRUE, row.names=1)
df_signif = fread(synGet("syn22891771")$path) 
df_signif = df_signif[tissue =='Brain_Frontal_Cortex_BA9', ]
# identical(rownames(df_gtex.beta), rownames(df_gtex.se))
# identical(rownames(df_gtex.lfsr), rownames(df_gtex.se))

# data.frame of GTEx results
df_gtex_BA9 = data.frame( ENSEMBL.full = rownames(df_gtex.beta),
                          ENSEMBL = trim_ensembl_ids( rownames(df_gtex.beta) ),
                          logFC = -1*df_gtex.beta[,'Brain_Frontal_Cortex_BA9'],
                          se = df_gtex.se[,'Brain_Frontal_Cortex_BA9'],
                          adj.P.Val = df_gtex.lfsr[,'Brain_Frontal_Cortex_BA9'])

df_gtex_BA9$isSignif = FALSE
df_gtex_BA9$isSignif[ df_gtex_BA9$ENSEMBL.full %in% df_signif$gene] = TRUE

df_gtex_BA9 = df_gtex_BA9 %>% getGeneSymbol( "ENSEMBL")

df_gtex_BA9$t = with(df_gtex_BA9, logFC / se)
df_gtex_BA9$P.Value = with(df_gtex_BA9, 2*pnorm(abs(t), lower.tail=FALSE))
# df_gtex_BA9$adj.P.Val = p.adjust(df_gtex_BA9$P.Value, "fdr")
df_gtex_BA9 = df_gtex_BA9[!is.na(df_gtex_BA9$logFC),]
rownames(df_gtex_BA9) = df_gtex_BA9$ENSEMBL

# data.frame of CMC results from meta-analysis
df = df_meta[['Reported_GenderMale']]
df$t = with(df, logFC / se)
rownames(df) = trim_ensembl_ids( rownames(df) )

# all genes
resConcord.gtex = plot_concordance( list('CMC v3' = df, 'GTEx BA9' = df_gtex_BA9), showGenes=df$Symbol[1:10] )

main = "Concordance with GTEx (all genes)"
plot_grid(  resConcord.gtex$fig_logFC + ggtitle(main),
            resConcord.gtex$fig_tstat + ggtitle(main), nrow=1)

# resConcord.gtex$tab_pi %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.gtex$tab_corr %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)

# because GTEx uses MASH to combine 44 tissue, standard plot_concordance() does give 
# correct replication values
df.merge = merge(df, df_gtex_BA9, by="row.names")

table(df.merge$isSignif)

df_repl = data.frame(Discovery = c("GTEx", "CMC"), rate = c(NA, NA))

# Discover in CMC
# with(df.merge, table(isSignif[adj.P.Val.x < 0.05]))
# with(df, 1 - qvalue(P.Value.y[adj.P.Val.x < 0.05])$pi0)
lfsr = with(df.merge, adj.P.Val.y[adj.P.Val.x < 0.05]) 
df_repl$rate[1] = sum(1-lfsr) / length(lfsr)

# Discover in GTEx
# with(df, 1 - qvalue(P.Value.x[adj.P.Val.y < 0.05])$pi0)
# df_repl$rate[2] = with(df.merge, 1 - qvalue(P.Value.x[isSignif])$pi0)

res = with(df.merge, ash( logFC.x, se.x))
df.merge$lsfr = res$result$lfsr

lfsr = with(df.merge, lsfr[adj.P.Val.x < 0.05]) 
df_repl$rate[2] = sum(1-lfsr) / length(lfsr)

df_repl %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)

# how any DEGs?
# sum(apply(df_gtex.lfsr, 1, function(x) any(x <= 0.05, na.rm=TRUE)))

# df_combine = apply(df_gtex.lfsr, 2, function(x) sum(x <= 0.05, na.rm=TRUE))
# df_combine = data.frame(Tissue = names(df_combine), Count = df_combine)
# df_combine$Tissue = factor(df_combine$Tissue, rev(df_combine$Tissue))

# ggplot(df_combine, aes(Tissue, Count)) + geom_bar(stat="identity") + theme_bw() + coord_flip()

# # All genes: 
# sum(df_gtex.lfsr[,'Brain_Frontal_Cortex_BA9'] < 0.05, na.rm=TRUE)
# sum(df_gtex.lfsr[,'Brain_Frontal_Cortex_BA9'] < 0.05, na.rm=TRUE)

# # expressed in CMC
# sum(df_gtex_BA9$adj.P.Val < 0.05, na.rm=TRUE)

# sum( df$adj.P.Val.y < 0.05)

# # get significant genes directly
# df_combine = df_signif[,data.frame(Count = length(gene)), by="tissue"]
# df_combine$tissue = factor(df_combine$tissue, rev(df_combine$tissue))

# ggplot(df_combine, aes(tissue, Count)) + geom_bar(stat="identity") + theme_bw() + coord_flip()

```{r gtex.sex.2, fig.height=6, fig.width=10}
# autosomes
df = df[df$Chrom %in% 1:22,]

resConcord.gtex = plot_concordance( list('CMC v3' = df, 'GTEx BA9' = df_gtex_BA9), showGenes=df$Symbol[1:10] )

main = "Concordance with GTEx (autosomes)"
plot_grid(  resConcord.gtex$fig_logFC + ggtitle(main),
            resConcord.gtex$fig_tstat + ggtitle(main), nrow=1)

# resConcord.gtex$tab_pi %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.gtex$tab_corr %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)

df.merge = merge(df, df_gtex_BA9, by="row.names")

table(df.merge$isSignif)

df_repl = data.frame(Discovery = c("GTEx", "CMC"), rate = c(NA, NA))

# Discover in CMC
# with(df.merge, table(isSignif[adj.P.Val.x < 0.05]))
# with(df, 1 - qvalue(P.Value.y[adj.P.Val.x < 0.05])$pi0)
lfsr = with(df.merge, adj.P.Val.y[adj.P.Val.x < 0.05]) 
df_repl$rate[1] = sum(1-lfsr) / length(lfsr)

# Discover in GTEx
# with(df, 1 - qvalue(P.Value.x[adj.P.Val.y < 0.05])$pi0)
# df_repl$rate[2] = with(df.merge, 1 - qvalue(P.Value.x[isSignif])$pi0)

res = with(df.merge, ash( logFC.x, se.x))
df.merge$lsfr = res$result$lfsr

lfsr = with(df.merge, lsfr[adj.P.Val.x < 0.05]) 
df_repl$rate[2] = sum(1-lfsr) / length(lfsr)

df_repl %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)
```

@@

```{r separate.test.sz}

df = merge(df_meta.separ[[1]], df_meta.separ[[2]], by="row.names")

df$tstat = with(df, (logFC.x - logFC.y) / sqrt(se.x^2 + se.y^2))

df$p = 2*pnorm(abs(df$tstat), 0, 1, lower.tail=FALSE)
df$FDR = p.adjust(df$p, "fdr")

head(df[df$FDR < 0.1,])

@@
